Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-11 paper.tex
Added more responses
more updates

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The top left hand corner
shows a structure a chemist would draw it -- every ``corner''
representing a carbon atom (which can also be represented as a C),
every O denoting oxygen, every N nitrogen, every H hydrogen, and so
on. This is the standard format chemists use when exchanging
information with their peers.  However, this format is not directly
amenable to data mining -- different atoms might be collapsed into one
node, and sets of atoms might have the same relevant properties
assigned. Hence, the top right hand corner of
Figure~\ref{figure:chemical-structures} shows an annotated graph
representation, taking into account that atoms, though being of
different elemental atom type, can have similar
properties -- \emph{e.g.}, both oxygen and nitrogen carry an ``On''
label in this graph.

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others). Although the
graph implicitly encodes the 3D structure of a molecule (when combined
with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure. In addition, the 3D structure may have
a geometrical arrangement of features (such as the presence of a
right-handed helix) which cannnot be encoded in the graph.

To extend the concept of three-dimensionality further, in many
biological systems the \emph{surface properties} of the molecule are
what appear to be relevant, as shown in the bottom right hand corner.
Hence, while graph mining on chemicals is certainly a promising avenue
to explore, one needs to keep in mind that not all molecular
properties are encoded in the graph (for example, neither rotational
states, nor different charge states are encoded) and thus often other
``descriptors'' need to be used to capture the essentials of a
chemical system \cite{bender2004}. Hence, one of the difficulties of a
cheminformatician is that the equivalence of particular atoms (or sets
of atoms) is usually not known from the onset but it rather depends on
the particular system considered -- involving a large amount of
trial-and-error in the choice of a ``proper'' description for
molecules in a particular case.

>>>>

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisified by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{wegner2006,horst2009}.

@@

Although graphs implicitly encodes the 3D structure of a molecule
(when combined with knowledge about favored bond angles and
distances), many different low energy 3D structures, or conformers,
may be consistent with the same graph structure. In addition, the 3D
structure may have a ``secondary'' geometrical arrangement of features
(such as the presence of a right-handed helix) which cannnot be
encoded in the graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity) that varies over the surface.

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Indeed, Fig.~\ref{figure:chemical-structures}
represents a hierarchy of representations. Thus, when searching for
substructures, a 2D representation is sufficient. When trying to
rationalize why a molecule binds to a protein, a 3D structure is
vital.

A key challenge in chemical data mining is the choice of
representation and how that can affect subsequent analysis
efforts. But even after one has been chosen, one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And is so,
what type of labels are required? Currently, many of these questions
are answered using trial and error.
